Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-26 paper-amia-summit-2019/manuscript.tex
initial final version by Alex

Specifically, we introduce and evaluate the effectiveness of distributed word representations (i.e. word embeddings) as well as lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning architectures, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

@@

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (\hl{an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table}~\ref{tab:datastat}). \\

@@

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} We skipped the words or punctuation marks, for which pre-trained embeddings could not be found. CRF utilized lexical features and POS tags.

\subsection*{\textit{Classifiers}}

@@

\textbf{Multi-Layer Perceptron}: MLP is a neural network, which consists of multiple fully connected layers map an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multi-layer perceptron with a single hidden layer. In MLP networks, there is no cycles or loops and information flows only forward, from the input layer through the hidden layer(s) to the output layer. This study utilized MLP with a nonlinear activation function (rectified linear unit) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we utilized dropout in fully connected layers by randomly hiding neurons during training. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.45\textwidth]{figures/mlp.eps}
    \caption{\textbf{A multilayer perceptron with a single hidden layer}}
    \label{fig:mlp}
\end{figure}

\textbf{Bidirectional Recurrent Neural Network}: BRNN is a neural network designed to capture sequential patterns by considering both past \& future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The output of BRNN layer is computed as an aggregation of outputs of the forward and backward RNNs. Gated Recurrent Units (GRU)\cite{chung2014empirical} capable of handling variable size input sequence and having internal memory, which can be reset, were utilized as an RNN. Figure~\ref{fig:crnn} represents the architecture of BRNN, in the case when a convolution layer is removed.

\textbf{Convolutional Recurrent Neural Network}: CRNN is a deep neural network architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, which consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input layer to produce a $m \times n_e$ matrix after fetching the pre-trained word vectors. \hl{This matrix represents embedding output, contain sufficient morpho-syntactic information for segmentation of email exchanges}. When POS tags are used in combination with word embeddings, \hl{POS tags are embedded with 10-dimensional POS vectors and concatenated with 300-dimensional word vectors to obtain new embedding} vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of convolution is to extract new features depending on the neighboring words. We used 1D convolution. In this layer, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ useful features are produced in the convolution layer for each word. A max pooling operation was then performed over time to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted in the convolution layer. RNN is usually used to capture long-range dependencies in a sequence of observations. Moreover, bidirectional RNNs are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of bidirectional RNN layer for classifying each word or punctuation into ``new segment'' and ``same segment'' classes. Since a fully connected layer has a greater number of parameters, they are more likely to excessively co-adapt with these parameters, which may cause over-fitting. To prevent this, we utilized dropout by randomly disregarding 50\% of the connections to the fully connected layer. After that, logistic sigmoid \hl{outputs the probability of each class}. We experimentally determined the optimal parameters and found that the best performance is achieved with 5-fold cross-validation when filter length in convolutional layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as a convolutional layer activation function, RNN layer activation is hyperbolic tangent and the number of recurrent units in RNN is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{\hl{source codes are available at https://github.com/teanalab/eCoaching-Text-Segmentation}}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of convolutional recurrent neural networks for automatic segmentation of e-Coaching emails}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds, \hl{in which each fold was used as a test set and remaining 4 folds was utilized as a train set}. We also estimate the area under \hl{the precision-recall curve (AUPR) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets.} \cite{davis2006relationship}

\section*{Results}
Our experiments span four dimensions. First, we report the optimal size of word embeddings and sliding window size of the MLP model. Second, results are reported with respect to ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes in Table~\ref{tab:result_base}. Third, classification performance of different machine learning methods are summarized in Table~\ref{tab:result_weighted_avg} when word embeddings or lexical features are used in combination with punctuation and POS features. Fourth, the impact of the individual features, as well as their combination on the segmentation task, is reported.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mlp-and-vector.eps}
    \caption{\textbf{Performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and corpus word embeddings with GloVe and word2vec models (left). MLP model on e-Coaching email segmentation by varying the sliding window size 2n (right).}}
    \label{fig:embedding-dimension-mlp}
\end{figure}

Figure~\ref{fig:embedding-dimension-mlp} (left) shows the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and corpus word embeddings with GloVe \footnote{https://nlp.stanford.edu/projects/glove/} and word2vec models. It was observed that best performance is achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300-dimensional word vectors for all deep learning models used in this study. For MLP models, the first layer input was prepared by the summation of the first $n$ word vectors in the sliding window and the last $n$ word vectors when a sliding window contains $2n$ words or punctuations. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of the MLP model on e-Coaching email segmentation by varying the size of the sliding window. It can be observed that the best performance of MLP is achieved when $n$ is 2. Therefore, results of MLP in the remaining experiments are reported with $n$ set to 2.

\begin{table}[ht]
\centering
\caption{\textbf{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used. The highest value for each performance metric is highlighted in bold.}}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Overall}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other models in terms of recall and F1-measure achieving 0.797 recall with 0.785 F1-measure for new segment detection. CRNN also shows superior performance in all performance metrics for weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-Measure. On the other hand, MLP has the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-Measure in new segment class and 0.984 F1-Measure overall, which corresponds to the second highest performance in identifying ``new segment'' as well as a weighted average over both classes. Experimental results indicate that performance of all classifiers as a weighted average over both classes is remarkably higher compared to ``new segment'' class, which is expected since 96.95\% of instances belong to the ``same segment'' class and 99.3\% of them are correctly classified. For example, as a weighted average over ``new segment'' and ``same segment'' classes, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure, respectively, compared to the ``new segment'' detection. \\

\begin{table}[ht]
\centering
\caption{\textbf{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when all features are used together. The highest value for each performance metric
is highlighted in bold.}}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Overall}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.817} & 0.710 & 0.760 & 0.986 & 0.987 & 0.986 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_weighted_avg} summarizes the results of all models for segmentation of e-Coaching emails when word embeddings or lexical features are used in combination with punctuation and POS features. Similar to results in Tables~\ref{tab:result_base}, CRNN demonstrates the highest performance among all methods achieving 0.864 recall with 0.825 F1-Measure for ``new segment'' detection and 0.990 precision with 0.989 recall and F1-Measure overall. BRNN and CRF show the lowest and second highest performance, respectively, for email segmentation among all methods. We observed that classification performance significantly improved for ``new segment'' class when word embeddings or lexical features are used in combination with punctuation and POS features. Precision increases by 3.96\%, -2.27\%, 12.71\% and 1.81\%; recall increases by 11.72\%, 19.73\%, 20.59\% and 8.41\%; and F1-measure increases by 8.05\%, 9.51\%, 16.22\% and 5.1\% for CRF, MLP, BRNN and CRNN methods, respectively, in new segment detection when all features are utilized together. Similarly, precision increases by 0.51\%, 0.41\%, 0.82\% and 0.41\%; recall increases by 0.41\%, 0.41\%, 0.72\% and 0.3\%; and F1-measure increases by 0.41\%, 0.41\%, 0.82\% and 0.3\% for CRF, MLP, BRNN and CRNN methods, respectively, in weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used in combination with punctuation and POS features.\\

\begin{table}[ht]
\centering
\caption{\textbf{\hl{Area under the precision-recall curve (AUPR) values of all classifiers demonstrating the impact of word embeddings, punctuation and POS features on e-Coaching email segmentation. Highest AUPR value for each feature set across all models is highlighted in boldface.}}}
\label{tab:result_aupr}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\textbf{Features} & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & 0.780 & 0.736 & 0.655 & \textbf{0.818} \\ \hline
 word embeddings + POS & 0.797 (+2.18\%) & 0.746 (+1.36\%) & 0.647 (-1.22\%) & \textbf{0.798 (-2.44\%)} \\ \hline
 word embeddings + punctuation & \textbf{0.876 (+12.31\%)} & 0.835 (+13.45\%) & 0.774 (+18.17\%) & 0.874 (+6.85\%) \\ \hline
 all features & \textbf{0.877 (+12.44\%)} & 0.842 (+14.4\%) & 0.770 (+17.56\%) & 0.867 (+6\%) \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_aupr} shows the impact of individual features as well as their combination on the segmentation task. \hl{Influence of the punctuation and POS features is also consistent in AUPR values, which increase by 12.44\%, 14.4\%, 17.56\% and 6\% for CRF, MLP, BRNN and CRNN methods, respectively, when all features are used together. Individually, POS and punctuation features also improve the performance of all classifiers except BRNN and CRNN when POS features are used. CRF achieved the highest AUPR when punctuation or all features are used together. On the other hand, CRNN demonstrates the highest AUPR when word embeddings are used individually or combined with POS features.}

>>>>

Specifically, we introduce and evaluate the effectiveness of distributed word representations (i.e. word embeddings) as well as lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

@@

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table~\ref{tab:datastat}). \\

@@

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings as lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} For words or punctuation marks, which do not have pre-trained embeddings, we utilized the embeddings of the same dimensionality trained on the experimental dataset. CRF utilized lexical features, POS tags and the preceding label.

\subsection*{\textit{Methods}}

@@

\textbf{Multi-Layer Perceptron}: MLP is a neural network, which consists of multiple fully connected layers that map an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} illustrates a multi-layer perceptron with a single hidden layer. MLPs have no cycles or loops. Information in them flows only forward, from the input layer through the hidden layer(s) to the output layer. The MLP in this study utilizes one hidden layer consisting of 128 neurons and rectified linear unit (ReLU) as a nonlinear activation function. In order to prevent over-fitting, we applied dropout (random masking of neurons) \cite{srivastava2014dropout} to fully connected layers during training. Dropout was also applied to fully connected layers in BRNN and CRNN.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.45\textwidth]{figures/mlp.eps}
    \caption{\textbf{Multi-layer perceptron with a single hidden layer}}
    \label{fig:mlp}
\end{figure}

\textbf{Bidirectional Recurrent Neural Network}: BRNN is a neural network designed to capture sequential patterns by considering both past  and future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The hidden state of BRNN is an aggregation of the hidden states of a forward and backward recurrent neural networks (RNNs). Gated Recurrent Units (GRU)\cite{chung2014empirical} capable of handling variable size input sequence and having internal memory, which can be reset, were utilized as an RNN. %Figure~\ref{fig:crnn} represents the architecture of BRNN, in the case when a convolution layer is removed.

\textbf{Convolutional Recurrent Neural Network}: CRNN \cite{treviso2017sentence} shown in Figure~\ref{fig:crnn} is a deep neural network architecture, which combines convolutional and recurrent layers. Our implementation of CRNN consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input and embedding layers to produce a $m \times n_e$ matrix after fetching the embeddings for words and punctuations in the input sequence. This matrix corresponds to the distributed representation of an input email exchange, which contains rich morpho-syntactic information that can be utilized for its segmentation. When POS tags are utilized along with word embeddings, they are represented with a 10-dimensional vector, which is concatenated with 300-dimensional word embeddings to obtain new embedding vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of a convolution layer is to extract new features for each word or punctuation mark based on the neighboring words or punctuation marks. A one-dimensional (1D) convolution operation is utilized in this layer in our implementation of BRNN. In 1D convolution, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ features are produced by the convolution layer for each word. A max pooling over time operation was then performed to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted by the convolution layer. Unidirectional RNNs are usually used to capture long-range dependencies in a sequence of observations. Bidirectional RNNs, on the other hand, are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of the bidirectional RNN layer for classifying each word or punctuation into ``new segment'' or ``same segment'' classes. Since a fully connected layer has a larger number of parameters, they are more likely to excessively co-adapt to  other parameters, which may result in over-fitting. To prevent this, we utilized dropout by randomly ignoring 50\% of the connections in the fully connected layer. Finally, logistic sigmoid outputs the probability of classifying or labeling each word or punctuation mark with ``same segment'' class. We experimentally determined the optimal parameters using 5-fold cross-validation and found out that the best performance is achieved when filter length in the convolution layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as an activation function in the convolution layer, hyperbolic tangent is used as an activation function in the bi-directional RNN layer and the number of dimensions in the hidden state of RNNs is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{source code of all methods is available at \url{https://github.com/teanalab/eCoaching-Text-Segmentation}}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of a convolutional recurrent neural network for automated segmentation of e-Coaching emails}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric, since it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5-fold cross-validation (one fold was used as a test set and the remaining 4 folds were utilized as a training set) and weighted macro-averaging over the folds. We also report the area under the precision-recall curve (AUPR), due to its effectiveness in measuring the performance of binary classifiers in the case of the datasets with imbalanced class distribution.\cite{davis2006relationship}

\section*{Results}
Our experiments spanned three dimensions. First, we determined the optimal size of word embedding vectors and the sliding window of MLP. Second, we evaluated the performance of different methods with respect to detecting ``new segment'' as well as the weighted average over ``new segment'' and ``same segment'' classes. Third, we assessed the impact of different types of features as well as their combination on the performance of different machine learning methods on the e-Coaching email segmentation task.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mlp-and-vector.eps}
    \caption{\textbf{F1-measure of CRNN on the task of e-Coaching email segmentation by varying the number of dimensions in pre-trained and corpus-based GloVe and word2vec embeddings (left). F1-measure of MLP on the task of e-Coaching email segmentation by varying the size of the sliding window (right).}}
    \label{fig:embedding-dimension-mlp}
\end{figure}

Figure~\ref{fig:embedding-dimension-mlp} (left) illustrates the performance of CRNN on the task of e-Coaching email segmentation by varying the number of dimensions in pre-trained and corpus-based GloVe \footnote{https://nlp.stanford.edu/projects/glove/} and word2vec embeddings. We observed that the best performance is achieved with pre-trained 300-dimensional word2vec word vectors, when three types of features are used together. Therefore, we report the results for other deep learning models used in this study when 300-dimensional word2vec embedding vectors are utilized. The input layer of MLP consists of a sum of embeddings of $n$ words or punctuation marks before and the sum of embeddings of $n$ words or punctuation marks after the word or punctuation mark, which is the center of a sliding window of $2n$ words or punctuation marks. Figure~\ref{fig:embedding-dimension-mlp} (right) demonstrates the performance of MLP on e-Coaching email segmentation by varying the size of the sliding window. It can be observed that the best performance of MLP is achieved when the size of the sliding window is 4 (or $n=2$). Therefore, MLP results in the remaining experiments are reported when $n$ is set to 2.

\begin{table}[ht]
\centering
\caption{\textbf{Performance of CRF, MLP, BRNN and CRNN on ``new segment'' detection as well as the weighted average over ``new segment'' and ``same segment'' classes when only lexical features are used. The highest value for each performance metric is highlighted in boldface.}}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Overall}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other methods in terms of recall and F1-measure achieving 0.797 recall and 0.785 F1-measure for new segment detection. CRNN also shows superior performance according to all performance metrics calculated as a weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-measure. On the other hand, MLP demonstrated the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-measure on new segment detection and 0.984 F1-measure overall, which corresponds to the second highest performance in identifying ``new segment''. CRF also demonstrated the second highest performance according to all metrics calculated as a weighted average over both classes. Experimental results indicate that the performance of all classifiers according to all metrics calculated as a weighted average over both classes is significantly higher than their performance on ``new segment'' detection, which is expected since 96.95\% of instances belong to the ``same segment'' class and 99.3\% of them are correctly classified. For example, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure calculated as a weighted average over ``new segment'' and ``same segment'' classes, compared to the ``new segment'' detection. \\

\begin{table}[ht]
\centering
\caption{\textbf{Performance of CRF, MLP, BRNN and CRNN on ``new segment'' detection as well as the weighted average over ``new segment'' and ``same segment'' classes when all types of features are used together. The highest value for each performance metric is highlighted in boldface.}}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Overall}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.817} & 0.710 & 0.760 & 0.986 & 0.987 & 0.986 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_weighted_avg} summarizes the results of all models on the task of segmentation of e-Coaching emails when word embeddings or lexical features are used in combination with punctuation and POS features. Similar to results in Tables~\ref{tab:result_base}, CRNN demonstrates the best performance among all methods achieving 0.864 recall with 0.825 F1-measure for ``new segment'' detection and 0.990 precision with 0.989 recall and F1-measure overall. BRNN and CRF demonstrated the lowest and second highest performance on the task of email segmentation among all methods, respectively. We observed that classification performance significantly improved for ``new segment'' detection when lexical features are used in combination with punctuation and POS features. Specifically, precision increases by 3.96\%, -2.27\%, 12.71\% and 1.81\%; recall increases by 11.72\%, 19.73\%, 20.59\% and 8.41\%; and F1-measure increases by 8.05\%, 9.51\%, 16.22\% and 5.1\% for CRF, MLP, BRNN and CRNN methods, respectively, on new segment detection when all types of features are utilized together. Similarly, precision increases by 0.51\%, 0.41\%, 0.82\% and 0.41\%; recall increases by 0.41\%, 0.41\%, 0.72\% and 0.3\%; and F1-measure increases by 0.41\%, 0.41\%, 0.82\% and 0.3\% for CRF, MLP, BRNN and CRNN methods, respectively, as a weighted average over ``new segment'' and ``same segment'' classes when lexical features are used in combination with punctuation and POS features.\\

\begin{table}[ht]
\centering
\caption{\textbf{Area under the precision-recall curve (AUPR) values of all classifiers demonstrating the impact of different types of features on e-Coaching email segmentation performance. Highest AUPR value for each feature set across all models is highlighted in boldface.}}
\label{tab:result_aupr}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\textbf{Features} & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & 0.780 & 0.736 & 0.655 & \textbf{0.818} \\ \hline
 word embeddings + POS & 0.797 (+2.18\%) & 0.746 (+1.36\%) & 0.647 (-1.22\%) & \textbf{0.798 (-2.44\%)} \\ \hline
 word embeddings + punctuation & \textbf{0.876 (+12.31\%)} & 0.835 (+13.45\%) & 0.774 (+18.17\%) & 0.874 (+6.85\%) \\ \hline
 all features & \textbf{0.877 (+12.44\%)} & 0.842 (+14.4\%) & 0.770 (+17.56\%) & 0.867 (+6\%) \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_aupr} shows the impact of different types of features as well as their combination on e-Coaching email segmentation performance. Punctuation and POS features have similar effect measured by the AUPR, which increases by 12.44\%, 14.4\%, 17.56\% and 6\% for CRF, MLP, BRNN and CRNN, respectively, when all features are used together. Individually, although punctuation features improve the performance of all classifiers, POS features improve the performance of only CRF and MLP. CRF achieved the highest AUPR when all types of features are used together. On the other hand, POS features degraded the AUPR of CRNN.
